The role of standing variation in geographic convergent adaptation

### 2014-09-29 standing-variation-paper.tex
changes throughout, also updated figs
abstract

\section*{Abstract}
The extent to which shared selection pressures led to a share genetic
response among populations touches on many questions in evolutionary
biology. For a number of well studied traits and species it appears
that convergent evolution within species is common. Here we explore
the role of standing genetic variation in the evolution of convergent
genetic responses in a geographically spread population. The
relatively slow geographic spread of selected alleles allows multiple
alleles to spread from standing and new mutation before any one comes
to dominate the population. When these alleles meet, their progress is
substantially slowed--if the alleles are selectively equivalent, and their initial patterning of the range can be
thought of as a random tessellation.
We extend our previous work on this topic 
by incorporating standing variation, and we show that various properties of a simple
model of geographic adaptation can be well understood by analogy to a Poisson
process models of crystallization. 
In this framework we derive the geographic scale over which a typical allele is expected
to dominate, and the time it takes the species to adapt as a whole. We
also study the proportion of adaptive alleles that arise from standing variation. Finally we explore how the negative
pleiotropic effects of alleles before an environment change can bias
the subset of alleles that get to contribute to a species adaptive response. 
Our results suggest that much of adaptation may be geographically
local even when selection pressures are wide-spread. We discuss the
implications of these results for arguments of species coherence and
the nature of divergence between species.

@@

Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism due to heterozygote advantage \citep[note that
the conditions for a balanced polymorphism are
complicated by the hemizygosity of males, see][]{hedrick2011population, Pamillo:1979}. 
Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, notably
those present in fava beans (although, It is unclear if this alone is enough of a
selection pressure to maintain the polymorphism in the face of these alleles
considerable malaria resistance advantage). 
The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
applies as well in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). 
Therefore, our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
We assume that before malaria became prevalent, 
G6PD deficiency alleles suffered a decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and hemizygote males. 
Assuming that the underlying causes and strength of this drop in fitness have not changed, 
we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
\geq 0.05$),
in order to have resulted in the equilibrium frequency seen today.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
\caption{ %
{\bf Characteristic length, as a function of selective disadvantage}, 
compared to the corresponding quantity without standing variation (equation \eqref{eqn:chi_new}) 
and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
The other parameters are chosen to match those of G6PD,
$\mu = 10^{-5}$ and $s_b = 0.05$. As the characteristic length with
only standing variation is independent of dispersal distance it is
plotted as a single black dotted line. 
} \label{Fig-G6PD-charlength}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_charlengths_sb}   %{charlen-by-sd-limit}
\caption{ %
{\bf Characteristic length, as a function of selective advantage},
holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-5}$.  
We compare this to the corresponding quantity without standing
variation (equation \eqref{eqn:chi_new}) 
and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
} \label{Fig-G6PD-charlength-sb}
\end{center}
\end{figure}

>>>>

% standing deleterious variation
% higher mutation rate => more importance of standing
% G6PD a combination of new and standing
% interpolates betw panmixia and 2010
% pleiotropy pre-shift biases adaptation
% establishment phase
% selective sweeps and species coherence, gene flow

\section*{Abstract}
The extent to which populations experiencing shared selective pressures
adapt through a shared genetic response
is relevant to many questions in evolutionary biology. 
In a number of well studied traits and species, 
it appears that convergent evolution within species is common. 
In this paper, 
we explore how standing, deleterious genetic variation contributes to
convergent genetic responses in a geographically spread population,
extending our previous work on the topic.
Geographically limited dispersal slows the spread of each selected allele,
hence allowing other alleles 
-- newly arisen mutants or present as standing variation -- 
to spread before any one comes to dominate the population. 
When such alleles meet, their progress is substantially slowed 
-- if the alleles are selectively equivalent, 
they mix slowly, dividing the species range into a random tessellation,
which can be well understood by analogy to a Poisson process model of crystallization. 
In this framework, 
we derive the geographic scale over which a typical allele is expected to dominate, 
the time it takes the species to adapt as a whole,
and the proportion of adaptive alleles that arise from standing variation.
Finally, we explore how negative pleiotropic effects of alleles before an environment change 
can bias the subset of alleles that get to contribute to a species adaptive response. 
We apply the results to the many geographically localized G6PD deficiency alleles thought to confer resistance to malaria,
whose large mutational target size and deleterious effects make them likely candidates to have been present as deleterious standing variation.
We find the numbers and geographic spread of these alleles matches our predictions reasonably well,
which suggest that these arose both from standing variation and new mutations since the advent of malaria.
Our results suggest that much of adaptation may be geographically local 
even when selection pressures are wide-spread. 
We close by discussing the implications of these results for arguments of species coherence and the nature of divergence between species.

@@

Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism due to heterozygote advantage \citep[note that
the conditions for a balanced polymorphism are
complicated by the hemizygosity of males, see][]{hedrick2011population, Pamillo:1979}. 
Indeed, hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, notably
those present in fava beans (although, It is unclear if this alone is enough of a
selection pressure to maintain the polymorphism in the face of these alleles
considerable malaria resistance advantage). 
The theory we use regarding the ``wave of advance'' \citep{fisher1937wave}
applies as well in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). 
Therefore, our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
We assume that before malaria became prevalent, 
G6PD deficiency alleles suffered a decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and hemizygote males. 
Assuming that the underlying causes and strength of this drop in fitness have not changed, 
we estimate that $s_d$ has to have been upward of $\sim 0.05$ (if $sb
\geq 0.05$),
in order to have resulted in the equilibrium frequency seen today in
areas with endemic malaria.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_charlengths}   %{charlen-by-sd-limit}
\caption{ %
{\bf Characteristic length, as a function of selective disadvantage}, 
compared to the corresponding quantity without standing variation (equation \eqref{eqn:chi_new}) 
and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
The other parameters are chosen to match those of G6PD,
$\mu = 10^{-5}$ and $s_b = 0.05$. As the characteristic length with
only standing variation is independent of dispersal distance it is
plotted as a single black dotted line. 
\plr{$y$ axis label: kilometers}
} \label{Fig-G6PD-charlength}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6PD_charlengths_sb}   %{charlen-by-sd-limit}
\caption{ %
{\bf Characteristic length, as a function of selective advantage},
holding the prior disadvantage of the allele fixed at $s_d=0.05$ and the mutation rate fixed at $\mu=10^{-5}$.  
We compare this to the corresponding quantity without standing
variation (equation \eqref{eqn:chi_new}) 
and to the quantity only considering standing variants (equation \eqref{eqn:chi_standing}).
\plr{$y$ axis label: kilometers}
} \label{Fig-G6PD-charlength-sb}
\end{center}
\end{figure}
